```{r}
library(data.table)
library(forecast)
library(fabletools)
library(arrow)
library(stringr)
library(plotly)
library(feasts)
library(lubridate)
library(ggplot2)
library(stringr)
library(dplyr)
library(tidyr)
library(purrr)
library(tsibble)
```

#load dataset
```{r}
table_names <- c(
  "BIDDAYOFFER_D", "BIDPEROFFER_D", "DEMAND", "INTERMITTENT_DS_RUN",
  "INTERMITTENT_DS_PRED", "INTERMITTENT_FORECAST_TRK", "OPERATIONAL_DEMAND",
  "ROOFTOP", "UNIT_SOLUTION", "CONSTRAINT", "OUTAGEDETAIL", "PUBLIC_PRICES"
)

# Load all datasets into a named list
datasets <- lapply(table_names, function(name) {
  open_dataset(file.path("filtered_parquet_tables", name))
})
names(datasets) <- table_names

# Access like this:
#datasets$BIDDAYOFFER_D
#datasets$ROOFTOP
datasets$DEMAND %>%collect() %>%distinct(DUID)

```

### head of each clean table

```{r}
# Preview first 6 rows from each table
walk(names(datasets), function(tbl) {
  cat("\n\n🔍 Table:", tbl, "\n")
  tryCatch({
    datasets[[tbl]] %>%
      collect() %>%
      head() %>%
      print()
  }, error = function(e) {
    cat("❌ Error collecting data for", tbl, ":", e$message, "\n")
  })
})

```

# EDA

## Cov analysis Demand(scada), operational demand, rooftop, public prices system-level(timestamp based)

```{r}
source("AggregationTime.R")
# ------------------ User Input: Region ------------------
# Use NULL to aggregate all regions
# Or use a vector of region IDs, e.g. region_filter <- c("VIC1", "NSW1")
region_filter <- NULL

# ------------------ Aggregation ------------------
bidper_df_cov       <- safe_agg_by_region_and_time(datasets$BIDPEROFFER_D, grep("BANDAVAIL", names(datasets$BIDPEROFFER_D), value = TRUE), region_filter)
bidday_df_cov       <- safe_agg_by_region_and_time(datasets$BIDDAYOFFER_D, grep("PRICEBAND", names(datasets$BIDDAYOFFER_D), value = TRUE), region_filter)
demand_df_cov      <- safe_agg_by_region_and_time(datasets$DEMAND, "SCADA_VALUE", region_filter)
op_demand_df_cov    <- safe_agg_by_region_and_time(datasets$OPERATIONAL_DEMAND, "OPERATIONAL_DEMAND", region_filter)
rooftop_df_cov      <- safe_agg_by_region_and_time(datasets$ROOFTOP, c("POWERMEAN", "POWERPOE50"), region_filter)
prices_df_cov       <- safe_agg_by_region_and_time(datasets$PUBLIC_PRICES, c("RRP", "TOTALDEMAND", "DEMANDFORECAST", "DISPATCHABLEGENERATION", "DISPATCHABLELOAD", "NETINTERCHANGE", "AVAILABLEGENERATION", "INITIALSUPPLY"), region_filter)
constraint_df_cov   <- safe_agg_by_region_and_time(datasets$CONSTRAINT, c("VIOLATIONDEGREE", "MARGINALVALUE", "RHS", "LHS"), region_filter)
in_ds_pred_df_cov   <- safe_agg_by_region_and_time(datasets$INTERMITTENT_DS_PRED, c("FORECAST_MEAN", "FORECAST_POE50"), region_filter)
unit_df_cov         <- safe_agg_by_region_and_time(datasets$UNIT_SOLUTION, c("AVAILABILITY", "TOTALCLEARED"), region_filter)

# ------------------ Join All Tables ------------------
system_region_df <- reduce(
  list(bidper_df_cov, bidday_df_cov, demand_df_cov, op_demand_df_cov, rooftop_df_cov, prices_df_cov, constraint_df_cov, in_ds_pred_df_cov, unit_df_cov),
  full_join,
  by = c("REGIONID", "timestamp_30m")
)

# Add residual demand
system_region_df <- system_region_df %>%
  mutate(residual_demand = OPERATIONAL_DEMAND - POWERMEAN)

system_scaled <- system_region_df %>%
  select(-REGIONID, -timestamp_30m) %>%
  scale()
```

```{r}
cov_matrix_region <- cov(system_scaled, use = "pairwise.complete.obs")
cor_matrix_region <- cor(system_scaled, use = "pairwise.complete.obs")

cov_matrix_region
cor_matrix_region
```

### per duid

```{r}
# =======================
# 🔁 Load Aggregation Function
# =======================
source("AggregationTime.R")  # Make sure this file contains your enhanced `safe_agg_by_duid_and_time()`

# =======================
# ⚙️ Variable Configuration
# =======================
aggregation_vars <- list(
  BIDDAYOFFER_D = grep("PRICEBAND", names(datasets$BIDDAYOFFER_D), value = TRUE),
  BIDPEROFFER_D = c(grep("BANDAVAIL", names(datasets$BIDPEROFFER_D), value = TRUE), "MAXAVAIL"),
  UNIT_SOLUTION = c("AVAILABILITY", "TOTALCLEARED"),
  DEMAND = "SCADA_VALUE",
  INTERMITTENT_DS_PRED = c("FORECAST_MEAN", "FORECAST_POE50"),
  CONSTRAINT = c("VIOLATIONDEGREE", "RHS", "LHS", "MARGINALVALUE"),
  PUBLIC_PRICES = c("RRP", "TOTALDEMAND", "DEMANDFORECAST", "DISPATCHABLEGENERATION",
                    "DISPATCHABLELOAD", "NETINTERCHANGE", "AVAILABLEGENERATION", "INITIALSUPPLY"),
  OPERATIONAL_DEMAND = "OPERATIONAL_DEMAND",
  ROOFTOP = c("POWERMEAN", "POWERPOE50")
)

# =======================
# 🧠 User Input
# =======================
duid_filter <- "LOYYB1"  # or NULL to aggregate across all DUIDs

# =======================
# 🔄 Aggregation Loop
# =======================
aggregated_list <- lapply(names(aggregation_vars), function(tbl_name) {
  if (!tbl_name %in% names(datasets)) return(NULL)
  
  vars <- aggregation_vars[[tbl_name]]
  df <- datasets[[tbl_name]]
  
  message("📊 Aggregating ", tbl_name)
  safe_agg_by_duid_and_time(df, vars, duid_filter)
})

# Remove any null results
aggregated_list <- aggregated_list[!sapply(aggregated_list, is.null)]

# =======================
# 🔗 Dynamic Join of Tables
# =======================

# Start with the first table
unit_level_df <- aggregated_list[[1]]

# Loop through the rest and join based on common columns
for (i in 2:length(aggregated_list)) {
  next_df <- aggregated_list[[i]]
  
  # Get common join keys (must include timestamp_30m)
  join_keys <- intersect(names(unit_level_df), names(next_df))
  join_keys <- join_keys[join_keys %in% c("REGIONID", "DUID", "timestamp_30m")]
  
  if (length(join_keys) < 1) {
    warning(paste("⚠️ Skipping join with dataset", i, "- no common join keys"))
    next
  }
  
  unit_level_df <- full_join(unit_level_df, next_df, by = join_keys)
}

# =======================
# ⚙️ Scale Data & Compute Covariance/Correlation
# =======================
unit_scaled <- unit_level_df %>%
  select(-REGIONID, -DUID, -timestamp_30m) %>%
  scale()
```

```{r}
head(unit_scaled)
head(unit_level_df)
# Run PCA
pca_result <- prcomp(unit_scaled, center = TRUE, scale. = TRUE)

# View variance explained
summary(pca_result)

# Scree plot
plot(pca_result, type = "lines")

# Biplot (first two PCs)
biplot(pca_result, scale = 0)

# Loadings (which variables influence each PC)
loadings <- pca_result$rotation

```

```{r}
cov_matrix_unit <- cov(unit_scaled, use = "pairwise.complete.obs")
cor_matrix_unit <- cor(unit_scaled, use = "pairwise.complete.obs")

# =======================
# 📤 Output
# =======================
print("📈 Covariance Matrix:")
print(cov_matrix_unit)

print("🔗 Correlation Matrix:")
print(cor_matrix_unit)

# (Optional) Save to CSV
# write.csv(cov_matrix_unit, "cov_matrix.csv")
# write.csv(cor_matrix_unit, "cor_matrix.csv")

```

##Visualisation 
### BID behaviour against other variable
adding RPP and demand to the supply plot

```{r}
source("BidBehaviour.R")
# Example usage
plot_supply_curve_animation(datasets$BIDDAYOFFER_D%>%collect(), datasets$BIDPEROFFER_D %>%collect(), "LOYYB1")
print(plot_band_avail_heatmap(datasets$BIDPEROFFER_D %>% collect(), "LOYYB1"))
```

#### RPP VS BID
```{r}
source("BidBehaviour.R")

plot_bid_vs_rrp_overlay_animated(
  bidday_df   = datasets$BIDDAYOFFER_D %>% collect(),
  bidper_df   = datasets$BIDPEROFFER_D %>% collect(),
  price_df    = datasets$PUBLIC_PRICES %>% collect(),
  duid_filter = "LOYYB1",                # Replace with your unit
  target_month = "2024-06"                   # NULL Or use "2024-06" for specific month
)
```
#### RPP vs BID vs RESIDUAL DEMAND (adding variabel based on the pca)
```{r}
# Source your cleaned data and function
source("BidBehaviour.R")

plot_bid_vs_rrp_overlay_with_residual_animated(
  bidday_df = datasets$BIDDAYOFFER_D %>% collect(),
  bidper_df = datasets$BIDPEROFFER_D %>% collect(),
  rrp_df    = datasets$PUBLIC_PRICES %>% collect(),
  op_demand_df = datasets$OPERATIONAL_DEMAND %>% collect(),
  rooftop_df = datasets$ROOFTOP %>% collect(),
  duid_filter = "LOYYB1",
  target_month = "2024-06",
  log_y = FALSE
)

```


```{r}
source("BidBehaviour.R")
# Option 1: Replace NA with 0
bid_tsibble <- prepare_bid_tsibble(datasets$BIDPEROFFER_D %>% collect(), "LOYYB1", impute_method = "zero")


# Plot with facets
plot_seasonal_decomp_facet(bid_tsibble)

```

### time series linear, log, sqrt (per variable)

```{r}
vic_ts <- system_region_df %>%
  filter(REGIONID == "VIC1") %>%
  select(timestamp_30m, OPERATIONAL_DEMAND, SCADA_VALUE, RRP, POWERMEAN, TOTALCLEARED) %>%
  as_tsibble(index = timestamp_30m)

vic_ts %>%
  pivot_longer(-timestamp_30m, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = timestamp_30m, y = value)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y")

vic_ts %>%
  mutate(across(where(is.numeric), list(
    sqrt = sqrt,
    log = ~log(. + 1)
  ))) %>%
  pivot_longer(-timestamp_30m, names_to = "transformation", values_to = "value") %>%
  ggplot(aes(x = timestamp_30m, y = value)) +
  geom_line() +
  facet_wrap(~transformation, scales = "free_y")
```

### Bid suppply curve, but only capture the lastest bidavail each day (perduid)

```{r}
plot_bids_paginated <- function(bidday_df, bidper_df, region = "VIC1", duid_filter = NULL, page = 1, page_size = 12) {
  
  # Step 1: Filter region & collect
  bidday_df_filtered <- bidday_df %>%
    filter(REGIONID == region) %>%
    collect()
  
  bidper_df_filtered <- bidper_df %>%
    filter(REGIONID == region) %>%
    collect()
  
  # Step 2: Latest bids per DUID
  latest_bidday <- bidday_df_filtered %>%
    group_by(DUID) %>%
    filter(timestamp == max(timestamp)) %>%
    ungroup()
  
  latest_bidper <- bidper_df_filtered %>%
    group_by(DUID) %>%
    filter(timestamp == max(timestamp)) %>%
    ungroup()
  
  # Step 3: Extract bands
  price_cols <- grep("PRICEBAND", names(latest_bidday), value = TRUE)
  avail_cols <- grep("BANDAVAIL", names(latest_bidper), value = TRUE)
  
  price_long <- latest_bidday %>%
    select(DUID, all_of(price_cols)) %>%
    pivot_longer(-DUID, names_to = "Band", values_to = "Price") %>%
    mutate(Band = as.integer(str_extract(Band, "\\d+")))
  
  avail_long <- latest_bidper %>%
    select(DUID, all_of(avail_cols)) %>%
    pivot_longer(-DUID, names_to = "Band", values_to = "MW") %>%
    mutate(Band = as.integer(str_extract(Band, "\\d+")))
  
  band_df <- left_join(price_long, avail_long, by = c("DUID", "Band"))
  
  # Step 4: Filter by DUIDs (optional)
  if (!is.null(duid_filter)) {
    band_df <- band_df %>% filter(DUID %in% duid_filter)
  }
  
  # Step 5: Pagination
  all_duids <- sort(unique(band_df$DUID))
  start_idx <- ((page - 1) * page_size + 1)
  end_idx <- min(start_idx + page_size - 1, length(all_duids))
  selected_duids <- all_duids[start_idx:end_idx]
  
  band_df <- band_df %>% filter(DUID %in% selected_duids)
  
  # Supply curve
  supply_df <- band_df %>%
    arrange(DUID, Price) %>%
    group_by(DUID) %>%
    mutate(CumMW = cumsum(MW)) %>%
    ungroup()
  
  # Plot 1 - Price
  p1 <- ggplot(band_df, aes(x = Band, y = Price, fill = DUID)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ DUID, scales = "free_y") +
    labs(title = paste("Offered Price - Page", page), x = "Band", y = "Price") +
    theme_minimal()
  
  # Plot 2 - Quantity
  p2 <- ggplot(band_df, aes(x = Band, y = MW, fill = DUID)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ DUID, scales = "free_y") +
    labs(title = "Offered MW", x = "Band", y = "MW") +
    theme_minimal()
  
  # Plot 3 - Supply Curve
  p3 <- ggplot(supply_df, aes(x = CumMW, y = Price, group = DUID)) +
    geom_step(aes(color = DUID), size = 1) +
    facet_wrap(~ DUID, scales = "free") +
    labs(title = "Supply Curve", x = "Cumulative MW", y = "Price") +
    theme_minimal()
  
  # Return interactive plots
  list(
    ggplotly(p1),
    ggplotly(p2),
    ggplotly(p3)
  )
}

```

```{r}
# Manually select DUIDs of interest
plots_specific <- plot_bids_paginated(
  bidday_df = datasets$BIDDAYOFFER_D,
  bidper_df = datasets$BIDPEROFFER_D,
  region = "VIC1"
  #duid_filter = c("LYB1", "OSY2")  # any DUIDs you want
)

plots_specific[[1]]
plots_specific[[2]]
plots_specific[[3]]  # View supply curve

```

```{r}


# Filter one unit
target_duid <- "LOYYB1"
bidper_df_graph4 <- datasets$BIDPEROFFER_D %>%
  filter(DUID == target_duid) %>%
  collect()

# Get all BANDAVAIL columns
avail_cols <- grep("BANDAVAIL", names(bidper_df), value = TRUE)

# Convert from wide to long format
long_bid <- bidper_df_graph4 %>%
  select(timestamp, all_of(avail_cols)) %>%
  pivot_longer(cols = starts_with("BANDAVAIL"), names_to = "Band", values_to = "MW") %>%
  mutate(
    Band = as.integer(str_extract(Band, "\\d+")),
    date = as_date(timestamp)
  )

# Daily average to reduce noise
long_bid_daily <- long_bid %>%
  group_by(Band, date) %>%
  summarise(MW = mean(MW, na.rm = TRUE), .groups = "drop")

# Convert to tsibble
bid_tsibble <- long_bid_daily %>%
  as_tsibble(index = date, key = Band) %>%
  fill_gaps(.full = TRUE)  # fills missing dates per band

# Plot seasonal pattern using gg_season
gg_season(bid_tsibble, MW) +
  labs(title = "Seasonality of Offered MW by Band", y = "MW") +
  theme_minimal() +
  facet_wrap(~ Band)

autoplot(bid_tsibble)
```

```{r}


# 1. Choose the DUID
target_duid <- "LOYYB1"

# 2. Load and filter BIDPEROFFER_D
bid_data <- datasets$BIDPEROFFER_D %>%
  filter(DUID == target_duid) %>%
  collect() %>%
  mutate(timestamp_5min = floor_date(timestamp, "5 minutes"))

# Combine all BANDAVAILXX columns into long format
bid_long <- bid_data %>%
  pivot_longer(
    cols = starts_with("BANDAVAIL"),
    names_to = "Band",
    values_to = "MW"
  )

# 3. Load and filter PUBLIC_PRICES (VIC1 only for now)
rrp_data <- datasets$PUBLIC_PRICES %>%
  filter(REGIONID == "VIC1") %>%
  collect() %>%
  mutate(timestamp_5min = floor_date(timestamp, "5 minutes")) %>%
  select(timestamp_5min, RRP)

# 4. Join bid + RRP
joined_df <- bid_long %>%
  left_join(rrp_data, by = "timestamp_5min")

# 5. Filter for valid numeric MW
joined_df <- joined_df %>% filter(!is.na(MW), !is.na(RRP))

# 6. Plot MW vs RRP per band
ggplot(joined_df, aes(x = RRP, y = MW)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~ Band, scales = "free_y") +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(
    title = paste("Bid MW vs RRP for", target_duid),
    subtitle = "Each band availability (MW) against 5-minute RRP",
    x = "RRP ($/MWh)", y = "Band MW"
  ) +
  theme_minimal()

```

```{r}
target_duid <- "LOYYB1"

# 1. Extract BIDPEROFFER data
bidper_df <- datasets$BIDPEROFFER_D %>%
  filter(DUID == target_duid) %>%
  collect() %>%
  mutate(timestamp_5min = floor_date(timestamp, "5 minutes")) %>%
  pivot_longer(
    cols = starts_with("BANDAVAIL"),
    names_to = "Band",
    values_to = "MW"
  ) %>%
  filter(!is.na(MW)) # remove NA band MWs

# 2. Extract PUBLIC_PRICES (e.g., VIC1)
rrp_df <- datasets$PUBLIC_PRICES %>%
  filter(REGIONID == "VIC1") %>%
  select(RRP, timestamp) %>%
  collect() %>%
  mutate(timestamp_5min = floor_date(timestamp, "5 minutes")) %>%
  distinct(timestamp_5min, .keep_all = TRUE) # Avoid duplicate RRP per interval

# 3. Join bid data with RRP
joined_df <- bidper_df %>%
  left_join(rrp_df, by = "timestamp_5min", relationship = "many-to-many") %>%
  filter(!is.na(RRP)) # clean post-join

# 4. Plot: Band MW vs RRP to detect correlation
ggplot(joined_df, aes(x = RRP, y = MW)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "#0077b6", se = FALSE) +
  facet_wrap(~ Band, scales = "free_y") +
  labs(
    title = paste("Reactive Bidding of", target_duid, "across Price Bands"),
    subtitle = "Does MW offered respond to changes in RRP?",
    x = "RRP ($/MWh)",
    y = "Offered MW (per Band)"
  ) +
  theme_minimal()

```

# not used

## Standardise time into timestamp variable (done in data procesing and cleaning)

```{r}
normalize_timestamp <- function(df, table_name) {
  df <- collect(df)
  
  if (table_name == "BIDDAYOFFER_D") {
    df$timestamp <- as.POSIXct(df$SETTLEMENTDATE, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "BIDPEROFFER_D") {
    df$timestamp <- as.POSIXct(df$SETTLEMENTDATE, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane") +
      lubridate::minutes((df$PERIODID - 1) * 5)
  } else if (table_name == "DEMAND") {
    df$timestamp <- as.POSIXct(df$RUN_DATETIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "INTERMITTENT_DS_RUN") {
    df$timestamp <- as.POSIXct(df$RUN_DATETIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "INTERMITTENT_FORECAST_TRK") {
    df$timestamp <- as.POSIXct(df$SETTLEMENTDATE, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "OPERATIONAL_DEMAND") {
    df$timestamp <- as.POSIXct(df$INTERVAL_DATETIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "ROOFTOP") {
    df$timestamp <- as.POSIXct(df$INTERVAL_DATETIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "UNIT_SOLUTION" || table_name == "CONSTRAINT") {
    date <- substr(df$DISPATCH_INTERVAL, 1, 8)
    period <- as.numeric(substr(df$DISPATCH_INTERVAL, 9, 11))
    df$timestamp <- as.POSIXct(date, format = "%Y%m%d", tz = "Australia/Brisbane") + lubridate::minutes((period - 1) * 5)
  } else if (table_name == "PUBLIC_PRICES") {
    df$timestamp <- as.POSIXct(df$SETTLEMENTDATE, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else if (table_name == "OUTAGEDETAIL") {
    df$actual_start <- as.POSIXct(df$ACTUAL_STARTTIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
    df$actual_end   <- as.POSIXct(df$ACTUAL_ENDTIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
  } else {
    warning("⚠️ No timestamp logic defined for table: ", table_name)
  }

  return(df)
}

```

```{r}
normalized_datasets <- lapply(names(datasets), function(name) {
  message("⏳ Normalizing: ", name)
  tryCatch({
    normalize_timestamp(datasets[[name]], name)
  }, error = function(e) {
    message("⚠️ Skipped ", name, ": ", e$message)
    NULL
  })
})
names(normalized_datasets) <- names(datasets)

```

## Normalised outgendetail alone

```{r}
library(arrow)
library(dplyr)
library(purrr)

process_outage_in_chunks <- function(dataset, chunk_size = 500000) {
  # Get total number of rows from metadata without full collect
  total_rows <- dataset %>% Scanner$create() %>% count_rows()
  num_chunks <- ceiling(total_rows / chunk_size)
  message("🔄 Processing ", total_rows, " rows in ", num_chunks, " chunks")

  
  # Initialize output list to save chunks
  result <- list()

  for (i in seq_len(num_chunks)) {
    message("⏳ Chunk ", i, " of ", num_chunks)

    start <- (i - 1) * chunk_size
    nrows <- min(chunk_size, total_rows - start)

    # Efficiently read just the slice
    chunk_df <- dataset %>%
      Scanner$create(skip = start, take = nrows) %>%
      collect()

    # Parse the timestamps
    chunk_df <- chunk_df %>%
      mutate(
        actual_start = as.POSIXct(ACTUAL_STARTTIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane"),
        actual_end   = as.POSIXct(ACTUAL_ENDTIME, format = "%Y/%m/%d %H:%M:%S", tz = "Australia/Brisbane")
      )

    result[[i]] <- chunk_df
  }

  bind_rows(result)
}

```

```{r}
normalized_outage <- process_outage_in_chunks(datasets[["OUTAGEDETAIL"]])

```

```{r}
normalized_datasets[["OUTAGEDETAIL"]] <- normalized_outage

```

## Data Visualisation

### Visualisaiton mean availibiltiy vs time

```{r}
# Get the top 10 DUIDs by mean availability
top_duids <- dispatchsumgraph %>%
  group_by(DUID) %>%
  summarise(total_availability = mean(mean_availability, na.rm = TRUE)) %>%
  arrange(desc(total_availability)) %>%
  slice_head(n = 10) %>%  # Select top 10
  pull(DUID)

# Filter dataset for only the top 10 DUIDs
dispatch_filtered <- dispatchsumgraph %>%
  filter(DUID %in% top_duids)

# Plot with better x-axis formatting
ggplot(dispatch_filtered, aes(x = HOUR, y = mean_availability, color = DUID)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +  # Reduce x-axis labels
  labs(title = "Hourly Availability Trends (Top 10 DUIDs)",
       x = "Time (Aggregated Hourly)",
       y = "Mean Availability (MW)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis for readability

```

```{r}
# Get the top 10 DUIDs by mean total cleared energy
top_duids_cleared <- dispatchsumgraph %>%
  group_by(DUID) %>%
  summarise(total_cleared = mean(mean_cleared, na.rm = TRUE)) %>%
  arrange(desc(total_cleared)) %>%
  slice_head(n = 10) %>%  # Select top 10
  pull(DUID)

# Filter dataset for only the top 10 DUIDs (Total Cleared)
dispatch_filtered_cleared <- dispatchsumgraph %>%
  filter(DUID %in% top_duids_cleared)

# Plot Hourly Total Cleared
ggplot(dispatch_filtered_cleared, aes(x = HOUR, y = mean_cleared, color = DUID)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +  # Reduce x-axis labels
  labs(title = "Hourly Total Cleared Energy (Top 10 DUIDs)",
       x = "Time (Aggregated Hourly)",
       y = "Mean Total Cleared (MW)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis for readability

```

```{r}
ggplot(dispatchsumgraph, aes(x = mean_availability, y = mean_cleared)) +
  geom_point(alpha = 0.6) +
  labs(title = "Availability vs. Total Cleared Energy (Hourly)",
       x = "Mean Availability (MW)",
       y = "Mean Total Cleared (MW)") +
  theme_minimal()

```

### Public price RPP vs unused energy

```{r}
dispatchsumgraph_joined <- dispatchsumgraph %>%
  left_join(publicprice, by = "HOUR")

#Unused Capacity vs RRP
ggplot(dispatchsumgraph_joined, aes(x = RRP, y = unused_capacity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Unused Capacity vs. Market Price (Hourly)",
       x = "Spot Price (RRP $/MWh)",
       y = "Unused Capacity (MW)") +
  theme_minimal()

```

```{r}
#Unused Capacity vs Availability (to show proportion)
ggplot(dispatchsumgraph_joined, aes(x = mean_availability, y = unused_capacity)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Unused vs Total Available Capacity (Hourly)",
       x = "Mean Availability (MW)",
       y = "Unused Capacity (MW)") +
  theme_minimal()

```

```{r}
#Add Price Color Gradient
ggplot(dispatchsumgraph_joined, aes(x = mean_availability, y = unused_capacity, color = RRP)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Unused Capacity vs. Availability, Colored by Price",
       x = "Mean Availability (MW)",
       y = "Unused Capacity (MW)",
       color = "RRP ($/MWh)") +
  theme_minimal()

```

### covariance analysis RRP with key variable

```{r}
cov_data <- dispatchsumgraph_joined %>%
  select(RRP, TOTALDEMAND, mean_cleared, mean_availability, unused_capacity)

cov_matrix <- cov(cov_data, use = "complete.obs")
print(cov_matrix)

cor_matrix <- cor(cov_data, use = "complete.obs")
print(cor_matrix)

library(reshape2)

melted_cov <- melt(cov_matrix)

ggplot(melted_cov, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Covariance Matrix: Factors Affecting RRP",
       x = "", y = "", fill = "Covariance") +
  theme_minimal()

```

#### Cov analysis explanation

this could be like this because need to use region an duid

or do normalision becasue each region unique, normalising region

```{r}
# Ensure publicprice has hourly aggregation
publicprice <- publicprice %>%
  mutate(HOUR = floor_date(SETTLEMENTDATE, unit = "hour"))

# Merge price data with dispatchsumgraph
dispatch_price <- left_join(dispatchsumgraph, publicprice, by = "HOUR")

# Plot availability vs. RRP
ggplot(dispatch_price, aes(x = RRP, y = mean_availability)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +  # Add trend line
  labs(title = "Energy Availability vs. Market Price (RRP)",
       x = "Market Price (RRP) in $/MWh",
       y = "Mean Availability (MW)") +
  theme_minimal()

```

```{r}
# Merge demand data with dispatch summary
dispatch_demand <- left_join(dispatchsumgraph, publicprice, by = "HOUR")

# Plot Demand vs. Total Cleared
ggplot(dispatch_demand, aes(x = TOTALDEMAND, y = mean_cleared)) +
  geom_point(alpha = 0.1, color = "black") +
  geom_density2d(color = "blue")  +
  labs(title = "Total Cleared Energy vs. Market Demand",
       x = "Market Demand (MW)",
       y = "Mean Total Cleared (MW)") +
  theme_minimal()

```

```{r}


# Convert to time series format
availability_ts <- ts(dispatchsumgraph$mean_availability, frequency = 24)

# Fit an ARIMA model
availability_forecast <- auto.arima(availability_ts)

# Forecast for next 7 days
forecast_result <- forecast(availability_forecast, h = 24*7)

# Plot the forecast
autoplot(forecast_result) +
  labs(title = "Forecasted Availability for Next 7 Days",
       x = "Time",
       y = "Forecasted Availability (MW)")

```

## Public Prices data set

drop off, could be 18-19 and have not releasing the net day 5 day interval. missing the 5 minutes interval

```{r}
hourly_demand_by_region <- publicprice %>%
  mutate(HOUR = floor_date(ymd_hms(SETTLEMENTDATE), unit = "hour")) %>%
  group_by(HOUR, REGIONID) %>%
  summarise(TOTAL_DEMAND_MW = sum(TOTALDEMAND, na.rm = TRUE)) %>%
  arrange(HOUR, REGIONID)

# View result
head(hourly_demand_by_region)

```

```{r}
ggplot(hourly_demand_by_region, aes(x = HOUR, y = TOTAL_DEMAND_MW, color = REGIONID)) +
  geom_line(size = 1) +
  labs(
    title = "Hourly Total Demand by Region",
    x = "Hour",
    y = "Total Demand (MW)",
    color = "Region"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```

## Bideperday and bidperoffer analysis

### long format

```{r}
# 1. Bid prices long format
bid_price_long <- biddayoffer %>%
  pivot_longer(
    cols = starts_with("PRICEBAND"),
    names_to = "BAND",
    values_to = "PRICE"
  ) %>%
  mutate(BAND = gsub("PRICEBAND", "BAND", BAND))

# 2. Availability long format
bid_avail_long <- bidperoffer %>%
  pivot_longer(
    cols = starts_with("BANDAVAIL"),
    names_to = "BAND",
    values_to = "AVAILABILITY"
  ) %>%
  mutate(BAND = gsub("BANDAVAIL", "BAND", BAND))

```

```{r}
ggplot(bid_price_long, aes(x = PRICE)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  labs(title = "Distribution of Bid Prices", x = "Price ($/MWh)", y = "Frequency") +
  theme_minimal()
```

```{r}
# Filter for one DUID
duid_plot <- "LYA2"  # LYA 2-4

bid_avail_long %>%
  filter(DUID == duid_plot) %>%
  ggplot(aes(x = as.POSIXct(INTERVAL_DATETIME), y = AVAILABILITY, color = BAND)) +
  geom_line() +
  labs(title = paste("Availibility perband perday for", duid_plot),
       x = "Time", y = "AVAILIBILITY") +
  theme_minimal()

```

```{r}
#Correct join: by DUID, BAND, and DATE(INTERVAL_DATETIME)
bid_combined <- bid_avail_long %>%
  left_join(bid_price_long, 
            by = c("DUID", "BAND", "SETTLEMENTDATE")) 

ggplot(bid_combined,
       aes(x = INTERVAL_DATETIME, y = AVAILABILITY, fill = BAND)) +
  geom_area(position = "stack") +
  labs(title = "Stacked Band Availability Over Time", x = "Time", y = "MW") +
  theme_minimal()

```

```{r}
ggplot(bid_combined,
       aes(x = PRICE, y = AVAILABILITY, color = BAND)) +
  geom_point(alpha = 0.6) +
  labs(title = "Price vs Availability per Band", x = "Price ($/MWh)", y = "MW") +
  theme_minimal()
```

```{r}
ggplot(bid_combined,
       aes(x = PRICE, y = AVAILABILITY)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Bid Price vs Availability", x = "Price", y = "MW") +
  theme_minimal()

```

### ANALYTCAL ANALYSIS

```{r}
cor(bid_combined$PRICE, bid_combined$AVAILABILITY, use = "complete.obs")

```

```{r}
bid_combined %>%
  group_by(DUID) %>%
  summarise(
    avg_price = mean(PRICE, na.rm = TRUE),
    avg_avail = mean(AVAILABILITY, na.rm = TRUE)
  )
```

### shiny app - dashboard

animated bid stack changes enable dynamic filtering handling very large dataset, donwloading and data cleaning - timeseries analysis

## combined analysis with dispatchsum and publicprice, also other dataset

## analyse bidding strategies

## give region for each duid

avalibility vs total cleared (difference that is not dispatched ) - shifting probably with price (high closer to one)-need to normalise. human error etc.

correlates with the price, problably will highlight the external factor, the corr maybe not will be showing 0.5 or 0.7

find the correlation the number

find other correlation that is 0.4 or 0.1 find the correlation just any so

residual demand (demand that is not meet by renewable, wind and solar ) higher residual demand (change to graph 3)

is residual demand effecting the price shifting.
