---
title: "AemoETL: Introduction & Quickstart"
author: "TYCIA DEWI"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    css: vignette.css 
    includes:
      after_body: scrollspy.js
vignette: >
  %\VignetteIndexEntry{AemoETL: Introduction & Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval = FALSE
)
```

# What is AemoETL?

**AemoETL** is an R package that provides:

-   A **download–extract–transform** (ETL) pipeline for Australian Energy Market Operator (AEMO) data

-   A set of **filtering rules** to reduce large AEMO tables to just the columns and rows you care about

-   Convenient output to **Parquet**

-   A lightweight **Shiny dashboard** (and helper run_app()) for exploring:

    -   Supply-curve animations

    -   Band-availability heatmaps

    -   Bidstack snapshots & animations

    -   Seasonality plots

## Installation

```{r}
# from CRAN
install.packages("AemoETL")

# or latest development version from GitHub:
#remotes::install_github("tdew0002/EnergyMarketAnalysis")

```

## 1. Download & Parquet Conversion

1.  **`main()`** (in `etl/script/DownloadAemo.R`)

```{r}
library(AemoETL)
main()

```

-   **Interactive menu**: choose AEMO Archive vs Current; drill down into subfolders until ZIP files appear.

-   **Download location**: is in raw_data in data folder

-   **Extraction**:

    -   Every ZIP is unpacked into raw_data in data folder
    -   Inner ZIPs are recursively extracted
    -   All CSVs are then parsed by `extract_id_tables()` into named chunks (e.g. “BIDPEROFFER_D”, “DREGION”)

-   **Parquet output**: will be in parquet_tables in data

-   **Cleanup**: original CSVs are deleted to save space.

-   **Mapping file**: dataset_table_mapping.rds in mapping files

## 2. Filter & Enrich

Once you have raw Parquet under parquet_tables folder under data folder, apply your filtering rules:

```{r}
library(AemoETL)
run_filtering()

```

**`run_filtering()`** (in **R/data_processing.R**)

-   Reads each raw parquet from data ==\> parquet_tables ==\> <TABLE_NAME>

-   Applies your per-table `$filter`, `$select`, and `$timestamp` rule (see **R/filtering_rules.R**)

-   Writes **filtered** Parquet under data ==\> filtered_parquet_tables ==\> <TABLE_NAME>

Maintains a `.processed_files.txt` in each folder so that already-filtered files are skipped automatically.

## 3. Inspecting the Data

You can read your filtered tables directly with **arrow**:

```{r}
library(arrow)
ds <- open_dataset("data/filtered_parquet_tables/BIDPEROFFER_D")
ds %>%
  filter(REGIONID == "VIC1") %>%
  collect() %>%
  head()

```

## 4. Launch Shiny Dashboard

AemoETL ships with a Shiny app . To launch it:

```{r}
run_app()

```

This will spin up a dashboard where you can:

-   Animate supply curves

-   View band-availability heatmaps

-   See bidstack snapshots & day-long animations

-   Explore daily seasonality of price, availability, RRP, demand, renewables
