---
title: "NEM Bidding Analysis Report"
author: "TYCIA DEWI"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    css: vignette.css 
    includes:
      after_body: scrollspy.js
vignette: >
  %\VignetteIndexEntry{NEM Bidding Analysis Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


# National Energy Market (NEM)

## Introduction

State Electricity Commission (SEC) of Victoria is a “government-owned renewable energy company,” producing electricity from wind and solar power, to then be supplied to Victorian households and businesses (SEC Victoria, 2025). The organisation was re-founded in 2024 by the Victorian government with an aim to provide affordable electricity of up to 4.5 gigawatts of power while also growing the renewable energy workforce by creating over 59,000 jobs (State Government of Victoria, 2025; State Government of Victoria, 2024).

SEC Victoria operates within the Australian National Electricity Market (NEM), a “wholesale electricity market and the physical power system” which facilitates the selling and transporting of electricity in the area in which it operates (AEMO, 2025; DCCEEW, 2024). NEM operates in five interconnected national jurisdictions in the east and southern coast of Australia, which are Victoria, New South Wales, Queensland, the Australian Capital Territory and Tasmania (AEMO, 2025).

The NEM is made up of 5 key categories of players: (1) Generation & Storage, (2) Transmission, (3) Distribution, (4) Retailing and (5) Market Operator (SEC, 2025). First, the Generation & Storage is responsible for generating and storing electricity, this includes solar and wind farms, and coal and gas power stations. Second, Transmission is responsible for transporting high voltage electricity from generators to demand centres. Third, Distribution is responsible for transporting medium to low voltage electricity from transmission networks to customers. Fourth, Retailing is responsible for the purchasing of electricity on behalf of customers and liaise with customers in regards to customer service and billing for energy provision. In this ecosystem, SEC is both a Generator and Retailer, or also referred to as Gentailer. Lastly, the Market Operator, in this case the Australian Energy Market Operator (AEMO) is responsible to manage the NEM, in regards to overseeing and facilitating supply and demand, collecting payments, organising the delegation of energy produced and payment for generators (SEC, 2025).

## How is NEM Regulated?

The NEM is operated as a gross pooled market, where the sum of all electricity generated is pooled into a single supply (SEC, 2025). To determine which electricity will be chosen to meet the demand in the market, each generator in NEM is required to provide their bidding availability and bidding price at which they are willing to dispatch their power; bidding availability is submitted every five minutes and bidding price is submitted every 24 hours (SEC, 2025). The bidding price is submitted as a range, from Band 1 to 10, that is incremental in price – a bidding availability/volume (in MW) will be associated with each of the price bands (in \$ / MWh). While the bidding prices cannot be changed within the 24 hour period, the quantity of available energy within each price band can be made every 5 minutes according to market conditions (Dungey et al., 2017). AEMO will then aggregate all generator bids and stack them from lowest to highest price, to then first fulfill the market demand for energy starting with the lowest price energy offered. Generators will only be compensated for the energy that are supplied to the central pool of electricity (Dungey et al., 2017). Thus, in the NEM bidding system there is no guarantee that all energy produced will be bought and utilised, as it is a competitive market (Dungey et al., 2017).

Therefore, to maximise efficiency and profit, an energy company must have a bidding strategy to gain competitive advantage to maximise the amount of energy dispatched to fulfill the market demand.

As an Energy Market Analyst Intern in SEC, my role came in as part of SEC’s attempt to improve their bidding strategy in the NEM. To do so, my project focuses on helping visualise the historic energy bidding behaviour of the whole NEM area. By doing so, patterns and anomalies in the market could be spotted more easily to be analysed further.

To do so, I have constructed a dashboard which showcases four different visualisations; (1) Supply curve, (2) Heatmap, (3) Bidstack and (4) Seasonal plot. The four different visualisations gives insight into the different generator’s bidding behaviours. This dashboard will then be used by the analysts in SEC Victoria to explore further patterns in the NEM to further determine their future bidding strategy.

# Methods

## NEM Bidding Variables

There are five categories of NEM Bidding Variables, that will be utilised.

### Price Bands (PRICEBAND_1…10)

Each generator is required to divide their offered capacity into up to 10 price bands, with each band specifying an incremental volume (MW) and associated price (\$/MWh). Band 1 is the lowest price offered and Band 10 is the highest price offered. This price band is offered once every 24 hours.

In determining the offered price and volume for each band, the generators will take into consideration factors including expected fuel costs, heat rates, startup costs, expected demand, and risk of non-dispatch, into their strategy (cite).

An evident pattern in price bands variable is that baseload bands from 1 to 3 are usually low in variance as fuel costs are relatively stable. As for mid to higher price bands from 4 to 10, there is a higher variance as availability in these bands may be adjusted more frequently each day based on short-term market conditions, such as when spot price – how much electricity is required to maintain balance in the power system – increases (AEMC, 2025). During unforeseen circumstances, such as scarcity, generators may also place exorbitant bids of up to \$16,600/MWh in Band 10 to avoid the provision of energy unless it is an emergency (SEC,2025).

### Availability Bands (BANDAVAIL_1…10 and MAXAVAIL)

At every 5-minute dispatch interval, each generator is required to provide the MW availability for each price band. The MW sum of BANDAVAIL_1 to BANDAVAIL_10 should be equal to the variable MAXAVAIL, which is the maximum physical capability of the generator.

However, there are forecasting constraints especially for wind and solar units, which may lead to the number of MAXAVAIL exceeding the forecasted output, FORECAST_POE50. In this case, the utilised effective availability chosen will always be the forecasted output, which is FORECAST_POE50, because it prevents generators from underdelivering, which maintains system reliability (AEMO, 2023). For example, if the MAXAVAIL is 100 MW but the forecasted output, FORECAST_POE50 is 70 MW, BANDAVAIL might list a sum of 100 MW across the 10 bands, however, 30 MW must be trimmed out of the highest band to ensure that the effective availability is equal to the forecasted output.

### Regional Reference Price (RRP) and Total Demand

Energy is dispatched every 5-minute interval, with the lowest priced bids often dispatched first until the market demand is fulfilled. The highest priced bid that was last dispatched to satisfy the actual demand is RRP; the intersection between aggregate supply stack and the demand extraction (DIT).

For total demand, DREGION dataset provides TOTALDEMAND variable, however it was not chosen as there can be minor variations due to embedded generation, losses and constraint imbalances. Thus, the sum of TOTALCLEARED dispatched across all DUIDs from the dataset UNIT_SOLUTION was utilised instead, to ensure more accurate reflection to the market supplier output.

It is important to note that the offered price could also be negatively priced, as low as -\$1,000 per MWh. This has been done historically as a bidding strategy, as generators are compensated with the market price rather than their offered price (AER, 2023). Thus, setting a negative price increases the likelihood of the generator’s capacity to be chosen and dispatched by AEMO, compared to higher-priced competitors (AER, 2023). Negative pricing has been more common with the rise of renewable energy, as wind and solar plants have negligible marginal costs and nearly zero startup or shutdown expenses. Thus, they can bid negatively to secure dispatch when weather conditions are favorable (AER, 2023).

### Fuel Type and Technology Classification

The following are categories of fuel types:

Coal–Black Coal: as a baseload generator, coal production has a high cost to start and cool down, thus, production cannot be stopped and there is a constant baseload or minimum amount of energy that the generator needs to dispatch (Hurn et al., 2016). For this reason, it is a relatively low-cost energy source.

Gas–OCGT (Open Cycle Gas Turbine): as a peaking, dynamic, quick-to-start generator type that could start and stop quickly, however it has a high cost and is less efficient than baseload plants (Breeze, 2021).

Gas–CCGT (Combined Cycle Gas Turbine): a mid-merit generator type that is active during daytime with high demand and shuts off during low demand. This fuel type combines both gas and steam turbine to produce electricity more efficiently, and is of a moderate cost (IPIECA, 2022).

Hydro: this fuel type is flexible as it is easily dispatchable -- being able to start and stop quickly – and may be able to set high prices during demand peaks.

Wind: an intermittent energy source as power produced depends on natural weather factors – when wind speeds are within turbine limits -- near zero marginal as it is almost free to run after installation.

Solar: an intermittent energy source as power produced depends on sunlight availability, zero marginal cost during daylight.

Battery Storage: fuel type which stores electricity from renewables or grid, and releases it when needed, such as Lithium-ion or flow batteries. It is highly dispatchable and often bidding at high prices to earn arbitrage, due to the low purchase price point.

### Dispatchable Unit Identifier & Region ID

Each generator is given a Dispatchable Unit Identifier (DUID) by AEMO for identification purposes. This is important as the bidding availability and price is done per unit (Wattclarity, 2025).

Each region in which the NEM operates, QLD, NSW, VIC, SA and TAS, are assigned a Region ID (Wattclarity, 2025).

### Residual Demand

Residual demand is the remaining electricity load that is remaining after accounting for renewable energy supply, and thus must be fulfilled by conventional generators, electricity imports or storage systems (Do et al., 2021).

## ETL Processing

As the Market Operator, AEMO is responsible for gathering data and presenting it to the public as open data, freely available on their website (AEMO, 2025b). The raw data for this project originates from AEMO’s secure FTP (or HTTP) archives, provided in fixed-width text or CSV files. Each zip folder downloaded contains multiple CSV files, in the form of a (e.g., BIDDAYOFFER_D_YYYYMMDD.csv, BIDPEROFFER_D_YYYYMMDD.csv, etc.). These data will then undergo the ETL process as follows:

1.  First, AEMO’s archive index will be scraped by using rvest (R), to show all available dataset folders such as, BIDDAYOFFER_D, BIDPEROFFER_D, etc.\
2.  Second, the dataset folders are opened and the date range is chosen for download. The script will then locate the corresponding zip file to the date, (e.g., BIDDAYOFFER_D_20240315.zip), which will then be downloaded and unzipped into a local folder (e.g., `data/raw_data/BIDDAYOFFER_D/20240315/`).\
3.  Third, multi-table parsing is conducted, as many CSVs contain “I”, -prefixed headers within a single file. By utilizing the helper function `extract_id_tables()`, the function reads all of the lines and locate header rows with lines beginning with “I,” and splits the content into chunks per table header, to then auto detect a sanitised table name (e.g., BIDDAYOFFER_D, BIDPEROFFER_D). The function `fread()` is then used to parse each data chunk into a data.table. Thus, resulting in one data.table per logical table, which is then appended (row‐bind) across all CSVs for that date.\
4.  Fourth, once each table’s data is collected for a date (e.g., all BIDDAYOFFER_D chunks for 20240315), it is saved as a Parquet file under `data/parquet_tables/BIDDAYOFFER_D/20240315.parquet` using Arrow’s `write_parquet()`. This is done as converting to Parquet provides high‐performance columnar storage, especially beneficial for large‐scale AEMO files, making it more structured and efficient for processing, and reduces memory footprint and I/O time (Vohra, 2016). This format is chosen as SEC Victoria stores their datasets in the Parquet format, which could be utilised for both R and Phyton.\
5.  Fifth, mapping and version control is implemented, the first time a dataset table is encountered, its “table name” will then be added to `mapping/dataset_table_mapping.rds`. This mapping ensures that subsequent downloads recognize existing tables and avoid reprocessing Parquet files that have been stored.\
6.  Sixth, once Parquet files are successfully written, the original CSVs are removed to save disk space.

This ETL pipeline is run for each new date and new datasets, to ensure that they are ready for analysis. In this project, it was run until all required table datasets BIDDAYOFFER_D, BIDPEROFFER_D, INTERMITTENT_DS_PRED, UNIT_SOLUTION, and DREGION, has been processed and ready to analyse.

## Shiny App

### Supply Curve

The Supply Curve shows, for a single generator, how its available volume is distributed across the 10 price bands at every 5-minute dispatch interval. The user is able to choose the generator to be shown based on its DUID.

First, filtering the two input tables for the selected DUID:

```{r}
bidday <- bidday_df   %>% filter(DUID == duid_filter)
bidper <- bidper_df   %>% filter(DUID == duid_filter)
```

`bidday_df` (BIDDAYOFFER_D) holds the PRICEBAND_1…10 columns (the 24-hour bid prices).\

`bidper_df` (BIDPEROFFER_D) holds the BANDAVAIL_1…10 columns (the actual 5-minute availabilities).

Second, the ten band columns need to be reshaped from wide to long:

```{r}
daily_price <- bidday %>%
  select(timestamp, PRICEBAND_1:PRICEBAND_10) %>%
  pivot_longer(-timestamp, names_to="Band", values_to="Price") %>%
  mutate(Band = as.integer(str_extract(Band, "\\d+")))

avail_data <- bidper %>%
  select(timestamp, BANDAVAIL_1:BANDAVAIL_10) %>%
  pivot_longer(-timestamp, names_to="Band", values_to="MW") %>%
  mutate(Band = as.integer(str_extract(Band, "\\d+")))

```

Converting each of the ten band columns into rows (timestamp, Band, Price) and (timestamp, Band, MW) makes it easy to join on both the time and band number.

Third, join and compute the cumulative supply:

```{r}
joined <- inner_join(daily_price, avail_data, by = c("timestamp","Band")) %>%
  group_by(timestamp) %>%
  arrange(Price) %>%
  mutate(CumMW = cumsum(MW)) %>%
  ungroup()

```

The price and quantity for each band are merged, at each dispatch interval. Moreover, the data is ordered by ascending Price then taking a running total over MW yields the classical “bid-stack” cumulative volume.

Fourth, plot and animate:

```{r}
ggplot(joined, aes(x = CumMW, y = Price, group = timestamp)) +
  geom_step(color = "#0077b6") +
  scale_y_log10() +
  labs(
    title = paste("Supply Curve for", duid_filter, "— {frame_time}"),
    x = "Cumulative MW",
    y = "Price ($/MWh)"
  ) +
  transition_time(timestamp) +
  ease_aes('linear') +
  theme_minimal()

```

`geom_step()` draws the step-function: each horizontal segment shows the volume at a given price band.\

Log‐scale on the y-axis (`scale_y_log10()`) makes wide price swings (including negative or very high bids) readable alongside baseload bands.\

`transition_time(timestamp)` animates through every 5-minute snapshot, so you see the curve shift as availability changes.

### Heatmap

The availability heatmap visualises how a single unit’s available volume is distributed across its ten price bands over time. The process was done as follows:

First, data extraction & filtering:

```{r}
bid_filtered <- bidper_df %>%
  filter(DUID == duid_filter)

```

Starting from the BIDPEROFFER_D table (`bidper_df`), which contains the 5-minute “real-time” availability for each price band (`BANDAVAIL_1…BANDAVAIL_10`). Filtering to `DUID == duid_filter` isolates the one generator you’ve selected in the dashboard.

Second, reshaping to long format:

```{r}
avail_cols <- grep("BANDAVAIL", names(bid_filtered), value = TRUE)

bid_long <- bid_filtered %>%
  select(timestamp, all_of(avail_cols)) %>%
  pivot_longer(-timestamp, names_to = "Band", values_to = "MW") %>%
  mutate(Band = as.integer(str_extract(Band, "\\d+")))

```

pivot_longer() stacks the ten BANDAVAIL_i columns into rows of (timestamp, Band, MW) so that Band can be treated as a categorical y-axis. The numeric band index is then extracted with str_extract(...) which ensures correct ordering.

Third, plotting with geom_tile():

```{r}
ggplot(bid_long, aes(
    x    = timestamp,
    y    = factor(Band),
    fill = MW
  )) +
  geom_tile() +
  labs(
    title = paste("Availability Heatmap for", duid_filter),
    x     = "Time",
    y     = "Price Band"
  ) +
  scale_fill_viridis_c(name = "MW Available") +
  theme_minimal()

```

geom_tile() draws one colored rectangle per (timestamp, Band), that shaded by the available MW. The shading utilises a viridis color scale to indicate the amount of available MW, which gives perceptually uniform contrast and is friendly for color‐vision deficiencies. Lastly, the Band is converted to a factor which preserves the natural order 1→10 on the y-axis.

### Bidstack

The Bidstack visualisation shows how the combined supply curve of every generator in a region evolves through each 5-minute dispatch interval, and where demand clears the stack. The point where demand is fulfilled is where the dotted line meets the graphs.

First, define time window & region of the stack:

```{r}
day0  <- ymd(date)
start <- as_datetime(day0)
end   <- start + days(1)
regs  <- if (tolower(region)=="all") NULL else strsplit(region, ",\\s*")[[1]]

```

By first building the 24-hour window from start to end, for the chosen date and convert the region string into a vector.

Second, to average daily price bands:

```{r}
daily_price <- open_dataset(bids_dir) %>%
  filter(timestamp >= start, timestamp < end) %>%
  filter(REGIONID %in% regs) %>%
  select(REGIONID, DUID, starts_with("PRICEBAND")) %>%
  collect() %>%
  group_by(REGIONID, DUID) %>%
  summarise(across(starts_with("PRICEBAND"), mean, na.rm=TRUE)) %>%
  pivot_longer(starts_with("PRICEBAND"), names_to="Band", values_to="Price") %>%
  mutate(Band = as.integer(str_extract(Band, "\\d+")))

```

The mean price of each generator’s 24-hour bid prices is calculated. It will then be pivoted into a long (Band, Price) table for later joins.

Third, selecting the best intermittent forecast:

```{r}
interm_tbl <- open_dataset(interm_dir) %>%
  filter(timestamp >= start, timestamp < end) %>%
  collect() %>%
  group_by(DUID, timestamp) %>%
  filter(
    FORECAST_PRIORITY == max(FORECAST_PRIORITY),
    OFFERDATETIME    == max(OFFERDATETIME)
  ) %>%
  slice_head(n=1) %>%
  select(DUID, FORECAST_POE50, timestamp)

```

From INTERMITTENT_DS_PRED, only the highest‐priority and most recent FORECAST_POE50 per generator and interval is kept in the dataset.

Fourth, trim 5-minute availabilities by forecast:

```{r}
avail_all <- open_dataset(per_dir) %>%
  filter(timestamp >= start, timestamp < end) %>%
  filter(REGIONID %in% regs) %>%
  select(REGIONID, DUID, timestamp, starts_with("BANDAVAIL"), MAXAVAIL) %>%
  collect() %>%
  pivot_longer(starts_with("BANDAVAIL"), names_to="Band", values_to="MW") %>%
  mutate(Band = as.integer(str_extract(Band, "\\d+"))) %>%
  left_join(interm_tbl, by=c("DUID","timestamp")) %>%
  mutate(
    effective_maxavail = if_else(!is.na(FORECAST_POE50),
                                 pmin(FORECAST_POE50, MAXAVAIL),
                                 MAXAVAIL)
  ) %>%
  group_by(REGIONID, DUID, timestamp) %>%
  mutate(
    totalMW   = sum(MW, na.rm=TRUE),
    over      = totalMW - effective_maxavail,
    maxBandNZ = max(Band[MW>0], na.rm=TRUE),
    MW        = if_else(over>0 & Band==maxBandNZ,
                        pmax(MW - over, 0),
                        MW)
  ) %>%
  ungroup()

```

Each band’s 5-minute availability is then joined with its forecast. If the sum of all bands exceeds the forecasted output (FORECAST_POE50), trim the excess from the highest‐price band to ensure total availability never surpasses the forecast.

Fifth, map fuel types:

```{r}
fuel_df <- open_dataset(units_dir) %>%
  select(DUID, FuelType) %>%
  distinct() %>%
  collect()

```

Pull each generator’s technology classification for color‐coding.

Sixth, building cumulative supply curve per timestamp:

```{r}
df_all <- daily_price %>%
  inner_join(avail_all, by=c("REGIONID","DUID","Band")) %>%
  left_join(fuel_df,  by="DUID") %>%
  filter(!is.na(Price), !is.na(MW)) %>%
  group_by(timestamp, FuelType, Price, Band) %>%
  summarise(MW = mean(MW, na.rm=TRUE)) %>%
  arrange(timestamp, Price) %>%
  group_by(timestamp) %>%
  mutate(CumVol = cumsum(MW)) %>%
  ungroup()

```

Merge price and (trimmed) availability, then compute the running total of MW at each price level for every interval.

Seventh, fetch cleared demand:

```{r}
demand_ts <- open_dataset(units_dir) %>%
  filter(timestamp >= start, timestamp < end) %>%
  filter(REGIONID %in% regs) %>%
  select(timestamp, TOTALCLEARED) %>%
  collect() %>%
  group_by(timestamp) %>%
  summarise(demand = sum(TOTALCLEARED), .groups="drop")

```

Summing TOTALCLEARED yields the actual volume dispatched; its maximum‐price band sets the Regional Reference Price (RRP).

Eighth, plot & animate the bidstack:

```{r}
ggplot(df_anim, aes(fill=FuelType)) +
  geom_rect(aes(
    xmin = CumVol - MW, xmax = CumVol,
    ymin = 0,         ymax = Price
  )) +
  geom_vline(aes(xintercept=demand),
             linetype="dashed", color="steelblue") +
  geom_text(data=label_df,
            aes(x=demand, y=price_cap, label=label_both),
            hjust=1.1, vjust=-0.5, size=3) +
  coord_cartesian(ylim=c(0,price_cap)) +
  scale_x_continuous("Cumulative MW", labels=number_format(scale=1e-3,suffix="K")) +
  scale_y_continuous("Price ($/MWh)", labels=dollar_format()) +
  labs(
    title    = paste0("Bidstack on ", date, " (", region, ")"),
    subtitle = "Time: {frame_time}", fill="Fuel Type"
  ) +
  theme_minimal(base_size=14) +
  transition_time(timestamp) +
  ease_aes("linear")

```

`geom_rect()` stacks colored blocks by fuel type, which shows how each technology contributes to total capacity at each price. A dashed vertical line marks actual demand; a text label shows the clearing price/volume.\
`transition_time(timestamp)` animates through each 5-minute interval, revealing intraday shifts in both capacity and clearing price.

### Seasonality Plot

The seasonality plot decomposes and visualises the time-of-day seasonality for key bid and market metrics over a user-specified date range. It does this by overlaying each day’s 30-minute trajectory (“spaghetti”) with a bold average line, making it easy to see both daily variability and the underlying repeating cycle. By aligning time of day and overlaying multiple days, diurnal cycle is visualised to show the pattern of variation in quantity over the course of a 24-hour period. Internally, the function proceeds as follows:

First, define the analysis window:

```{r}
dt0 <- ymd(start_date, tz="Australia/Brisbane")
dt1 <- ymd(end_date,   tz="Australia/Brisbane") + days(1) - seconds(1)
floor30 <- function(x) floor_date(x, "30 minutes")

```

Builds a POSIX window from start_date to end of end_date and defines a helper to round every timestamp down to its enclosing 30-minute bin.

Second, compute daily average offer price:

Read BIDDAYOFFER_D, filter by region & window.\
Pivot the ten PRICEBAND\_ columns long, then floor each timestamp to 30-minute bins.\
Group by calendar date and compute the mean price per bin.\
Result: one line per day in price_df, later plotted with grey lines + bold steel-blue average.

Third, aggregate total availability per 30-minute:

Read BIDPEROFFER_D, filter by region & window.\
Sum across all BANDAVAIL\_ columns to get total MW available in each interval.\
Group by date and bin to produce a spaghetti of daily availability curves (avail_df), plus a red mean line.

Fourth, fetch RRP, residual demand, renewables & operational demand:

-   RRP: pull every 5-minute RRP from DREGION, assign date and bin.

-   Operational demand: same binning for OPERATIONAL_DEMAND.

-   Renewables: from UNIT_SOLUTION, filter FuelType “Solar”/“Wind”, sum TOTALCLEARED per bin.

-   Residual demand: merge demand & renewables per bin, subtract to get Residual = Demand – Renewable_MW.

Fifth, plot each metric:

For each of the six outputs—price, availability, RRP, residual, renewable generation, operational demand—the code uses:

```{r}
ggplot(df, aes(
  x     = as_hms(ts - floor_date(ts, "day")),
  y     = value,
  group = date
)) +
  geom_line(color="grey80", linewidth=0.4) +
  stat_summary(aes(group=1),
               fun       = mean,
               geom      = "line",
               linewidth = 1.1) +
  scale_x_time(..., labels=label_time("%H:%M")) +
  scale_y_continuous(...) +
  labs(
    title    = sprintf("Daily %s — %s (%s to %s)", metric_name, region, start_date, end_date),
    subtitle = "Grey = each day; <color> = average"
  ) +
  theme_minimal()

```

Grey spaghetti: shows each calendar day’s 30-minute profile.\
Bold colored line: the mean across days, highlighting the recurring pattern.

# Results

The dashboard is able to provide insight into generator bidding behaviour via the four visualisations.

### Supply Curve

```{r}

```

The supply curve can show how price and volume interact, by animating or plotting a generator’s supply curve—cumulative offered volume (MW) on the x-axis versus offer price (\$/MWh) on the y-axis—the dashboard makes visible how each tranche of capacity is “stacked” by price.

This visualisation helps indicate where the bulk of capacity lies, such as where a coal plant’s large low price block versus a Gas-OCGT peaker’s small high-price slice. Moreover, it also showcases intraday strategy shifts and how each generator adjusts its availability offering per band each day. Thus, helping analysts understand each unit’s cost structure and how strategic bidding behaviour evolves through the day.

### Availability Heatmap

```{r}

```

The availability heatmap aids in spotting unusual or strategic bidding availability, as the user is able to choose a generator by DUID and a date range, to showcase the MW offered in each band over time. The date range is shown on the x-axis and the different bands on the y-axis, colour-scaled by MW.

This visualisation showcases routine patterns in each generator’s bidding strategy, as to where they may have consistent gaps of availability or unavailability at the same hour each day. Moreover, it makes it easier to spot anomalies in bidding strategies as well, such as when there is an unplanned outage which leads to a generator withdrawing large capacity from a band.

### Bidstack Visuals

```{r}

```

The bidstack visuals aggregates all generators’ trimmed, banded offers into a single step-plot for a region at a given interval, coloured by fuel type. Thus, allowing users to compare bidding stacks across time, for example 8 am vs. 6 pm, and regions such as Queensland vs. Victoria. By doing so, regional differences in fuel consumption can be highlighted, where one region may consume higher levels of renewable energy compared to another. Therefore, allowing analysts to assess inter-regional constraints, and the progress of renewable energy penetrating the NEM.

### Seasonality Plot

```{r}

```

The seasonality plot overlay multiple days’ worth of a metric (e.g., residual demand or average price) by hour of day, often with a highlighted multi-day average. Showcasing diurnal cycles throughout the day, differences between weekday and weekend bidding behaviour or demand, as well as impact on external events.

By “flattening” time onto a 24-hour cycle, these plots help stakeholders pinpoint operational patterns and evaluate whether bidding strategies align with expected seasonal trends.

## Limitation & Improvements

Several measures could be taken for future improvements, including making the dashboard easily extendable and integrating the auto-download and processing script directly into the dashboard.

First, to make the dashboard easily extendable, a modular architecture could be adopted, with each visualization encapsulated in its own module or Shiny module. Moreover, clear interfaces could be defined further for adding new plots or data sources, and template modules and documentation could be provided to ensure that future team members or collaborators can add in new analyses (e.g., forecast accuracy plots, cost-benefit analyses) without having to rewrite the core code.

Second, to integrating the auto-download and processing script into the dashboard, the ETL pipeline could be embedded as a background R process or scheduled task within the Shiny server (using future/promises or a cron-style scheduler). Furthermore, exposing a “Refresh Data” button in the UI to trigger incremental downloads, Parquet conversion, and mapping updates on demand will help increase efficiency. Lastly, implementing progress indicators and logging in to inform users when new data are fetched, processed, and ready for visualization– will deliver an end-to-end, one-click solution.

These enhancements will transform the dashboard from a static set of analyses into a dynamic, maintainable platform—one that can grow organically as new data, metrics, or market insights emerge.

# References

-   Dungey et al. 2017 CESifo Working Paper no. 6819 Strategic Bidding of Electric Power Generating Companies: Evidence from the Australian National Energy Market
-   SEC Victoria, 2025 <https://www.secvictoria.com.au/?utm_source=google&utm_medium=sem&utm_campaign=poweringvictoria24&utm_content=generic&gad_source=1&gad_campaignid=22404807562&gclid=CjwKCAjw3f_BBhAPEiwAaA3K5MKDN_qn5U7Yt_T31C3W6WzOnJiD1XXXNxrRGmWqIXmfgs6Q5mth6RoCdCAQAvD_BwE>
-   State Government of Victoria, 2025 Government-owned renewable energy is leading the way \| vic.gov.au
-   State Government of Victoria, 2024 The SEC Is Back For Good \| Premier
-   AEMO, 2025 AEMO \| National Electricity Market (NEM)
-   DCCEEW, 2024 National Electricity Market - DCCEEW
-   AEMO 2023 <https://aemo.com.au/-/media/files/electricity/nem/planning_and_forecasting/solar-and-wind/enablement-of-bid-max-avail-for-semi-scheduled-generators/australian-wind-energy-forecasting-system-awefs-and-australian-solar-energy-forecasting-system-asefs.pdf>
-   AER 2023 State of the energy market 2023
-   AEMC 2025 <https://www.aemc.gov.au/energy-system/electricity/electricity-market/spot-and-contract-markets#:~:text=The%20spot%20price%20tells%20generators,kWh%20and%209.8c/kWh>.
-   A. Stan Hurn, Annastiina Silvennoinen, Timo Teräsvirta <https://doi.org/10.1002/jae.2452> CESifo Working Paper no. 6819 found in here
-   Breeze, P (2021) <https://www.sciencedirect.com/topics/engineering/open-cycle-gas-turbine>
-   IPIECA, 2022 <https://www.ipieca.org/resources/energy-efficiency-compendium/combined-cycle-gas-turbines-2022>
-   Wattclarity, 2025 [https://wattclarity.com.au/other-resources/glossary/#:\~:text=Dispatchable%20Unit%20ID%20(DUID);,(and%20some%20registered%20loads)](https://wattclarity.com.au/other-resources/glossary/#:~:text=Dispatchable%20Unit%20ID%20(DUID);,(and%20some%20registered%20loads)).
-   Do et al., 2021 <https://www.sciencedirect.com/science/article/pii/S0306261920316846>
-   AEMO, 2025b AEMO \| Nemweb data
-   Vohra, 2016 Apache Parquet \| SpringerLink
