---
title: "AemoETL: Shiny Dashboard Tour"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    css: vignette.css 
    includes:
      after_body: scrollspy.js
vignette: >
  %\VignetteIndexEntry{NEM Bidding Analysis Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval = FALSE
)

```

## Quickstart with Pre-filtered Data

If you’d rather *not* wait for the full ETL to run, you can grab a snapshot of our latest **data folder** from Google Drive:

[👉 Download data folder (ZIP)](https://drive.google.com/drive/folders/18xc-82ECxCGiL95a0BfP1pnXpiqm67bb?usp=sharing)

-   Download the **data** folder

-   Put that **data folder** into the **package root**

-   Now you should have additional **data** **folder** that includes **raw_data, parquete_tables and filtered_parquet_tables** your root that includes

-   The app will pick it up instantly. Once you have those files, launch the dashboard:

```{r}
library(AemoETL)
run_app()   # launches the app; stop it to continue reading
```

If you ever need to pull new or incremental data, just run the full ETL pipeline:

```{r}
AemoETL::main()       # download & convert
AemoETL::run_filtering()  # apply your filters

```

# Dashboard Structure

The app is organized into four main tabs. The left‐hand sidebar lets you switch among them; controls appear in the top‐left of each panel, plots on the right.

## 1. Supply Curve

-   **Controls** (left box):

    -   **DUID(s)**: choose one or more generating units

    -   **Month**: select any month, YYYY-MM format

    -   **Start**: click to animate

-   **Plot** (right): an **animated** step‐curve showing cumulative offer volume vs price over each 5-min interval in the month.

-   Grey background lines are each day’s curve; default colors highlight the unit you selected.

## 2. Heatmap

-   **Controls**:

    -   **DUID**: pick a single unit

    -   **Start**: render the heatmap

-   **Plot**: a **tile plot** with

    -   **x = timestamp** (5-min steps)

    -   **y = Bid BAND index**

    -   **fill = available MW**

-   Perfect for spotting when and at which bands each unit was active.

## 3. Bidstack

Two sub‐modes:

1.  **Static snapshot**

-   **Controls**:

    -   **Region** (e.g. VIC1, NSW1…)

    -   **Timestamp** (5-min slice)

    -   **Start Static**

-   **Plot**: a single cumulative supply curve at that exact 5-min point, colored by fuel type and dashed line marking total cleared.

2.  **Day-long animation**

-   **Controls**:

    -   **Region**

    -   **Date** (pick a day)

    -   **Start Animated**

-   **Plot**: for each 5-min time step of the day, shows the evolving bidstack and annotates marginal price & cleared volume.

## 4. Seasonality

-   **Controls**:

    -   **Region**

    -   **Date range** (start & end)

    -   **Start**

-   **Plots**: a tabbed box with six panels, each a “spaghetti + average” chart:

1.  **Price** — daily average of all PRICEBAND value

2.  **Availability** — sum of all BANDAVAIL per 30-min, gray = each day, red = average

3.  **RRP** — regional wholesale spot price, gray = each day, green = average

4.  **Residual Demand** — operational demand minus renewables, gray = each day, orange = average

5.  **Operational Demand** — meter-level demand, gray = each day, purple = average

6.  **Renewable (Solar + Wind)** — sum of cleared for solar & wind units, gray = daily curves, gold = average \`\`\`

-   Hover your mouse over any line for an interactive tooltip (with exact MW, price, timestamp, etc).

## Tips & Tricks

-   **Region vs DUID**: most plots filter by **REGIONID**, only the supply‐curve uses DUID selectors.

-   **Timezones**: all timestamps are in **Australia/Brisbane**; slider controls convert to UTC internally.

-   **Performance**: the app pre‐opens each Parquet directory with `arrow::open_dataset()`
